Finding Concurrency Bugs in Production

### 2020-05-30 src/paper.tex
Add new section and shorten the introduction

% TODO: Correctly align the caption with the listing frame
\usepackage{caption}
\DeclareCaptionFont{white}{\color{white}}
\DeclareCaptionFormat{listing}{\colorbox{darkgray}{\parbox{\dimexpr\linewidth-0.7em\relax}{#1#2#3}}}
\captionsetup[lstlisting]{format=listing,labelfont=white,textfont=white}

@@

\lstdefinestyle{mystyle}{
    backgroundcolor=\color{backcolour},
    commentstyle=\color{codegreen},
    keywordstyle=\color{magenta},
    numberstyle=\tiny\color{codegray},
    stringstyle=\color{codepurple},
    basicstyle=\ttfamily\footnotesize,
    breakatwhitespace=false,
    breaklines=true,
    captionpos=t,
    keepspaces=true,
    numbers=left,
    numbersep=5pt,
    showspaces=false,
    showstringspaces=false,
    showtabs=false,
    tabsize=2,
    frame=single,
    xleftmargin=2em,
    framexleftmargin=1.57em,
    framexrightmargin=-0.5em
}

\lstset{style=mystyle}
\renewcommand{\lstlistingname}{Code}
\Crefname{listing}{Code}{Codes}
\crefname{listing}{code}{codes}

@@

\begin{abstract}
% TODO: Why in production?
Today's computer architectures rely more and more on parallel computing.
Single core processors were widely replaced by multi-core architectures optimized to run processes and threads concurrently.
To utilize the full performance of the given hardware, programmers need to write their software with a high degree of parallelism.
This introduces the potential for bugs that would not occur in sequential programs due to non-determinism in the order of execution.
Such bugs can be very challenging to fix because they are not easily reproducible.
For this purpose tools and languages have been developed to ease the creation of concurrency-aware programs and to help programmers find concurrency bugs effectively.
This paper will evaluate three different techniques and concrete tool implementations of them, that are available to find, reproduce and fix concurrency bugs from the viewpoint of how usable they are in a production environment.
The requirements for these tools are that they need to be easy to deploy, they should bring only a minimal computational and storage overhead, they should have a high coverage with minimal false-positive reports and they need to enable the developer to quickly find the cause of a bug.
The techniques evaluated are: Dynamic code analysis with the record and replay tool rr and the data race detector ThreadSanitizer.
Concurrency-aware testing with a combined approach of evaluating thread schedules with delta-debugging to automatically pinpoint concurrency bugs.
Static code analysis with the methods of sequentialization and model-checking.
All tools are evaluated for the Go programming language but the concepts are mainly applicable to every other language as well.
\end{abstract}

@@

Writing concurrent programs can be very challenging.
Computational work gets distributed among different processes and threads that need to synchronize to be able to speed up execution time on multi-core architectures.
Due to this complexity it is no surprise that bugs that are correlated to concurrency and parallel program execution occur quite often.
These bugs are also the most difficult to debug because they are so complex in their nature and not easy to reproduce.~\cite{tu2019go}
This is because of the non-determinism that is introduced when multiple threads or processes are executed simultaneously.
In contrast to sequential programs that only run on one thread and process, concurrent programs have multiple paths of execution that are executed more or less independently from each other.
The order these threads are executed is determined by the operating system's scheduler.
Because a developer, like the program itself, cannot predict what resources also need processing time on a machine, the concrete schedule has to be handled as non-predictable and therefore as non-deterministic.
But non-deterministic behavior also means that a bug that is only triggered when a certain pattern of thread executions or thread interleavings appear, cannot be reliably reproduced without knowing the schedule it appeared in.

@@

\begin{quote}
``The assertion that most production software bugs are soft -- Heisenbugs that go away when you look at them -- is well known to systems programmers. Bohrbugs, like the Bohr atom, are solid, easily detected by standard techniques, and hence boring. But Heisenbugs may elude a bugcatcher for years of execution. Indeed, the bugcatcher may perturb the situation just enough to make the Heisenbug disappear. This is analogous to the Heisenberg Uncertainty Principle in Physics.''\cite{gray1986computers}
\end{quote}

The complexity and uncertainty of concurrency bugs is also the reason for many studies that examine how these bugs occur and explore different ways on how to detect, fix or even avoid them.~\cite{tu2019go}
Thus, numerous tools and techniques have been developed that propose different solutions to solve the problem of concurrency bugs and accelerate the debugging process.

@@

\subsection{The Go Programming Language}
All code examples are written in Go which is a statically typed programming language developed by Google, which compiles down to single native binaries.
The reason we are using Go and not C/C++ or Java is the recent popularity of it for writing highly concurrent programs.
Popular open-source projects like \emph{Kubernetes}~\footnote{\url{https://kubernetes.io/}}, \emph{Docker}~\footnote{\url{https://www.docker.com/}} or \emph{Prometheus}~\footnote{\url{https://prometheus.io/}} are all written in Go.
In the official documentation of Go it claims: ``Go is expressive, concise, clean, and efficient. Its concurrency mechanisms make it easy to write programs that get the most out of multicore and networked machines, while its novel type system enables flexible and modular program construction.''~\cite{goDocs}
So, Go tries to solve some of the problems that arise when writing concurrent software application by their design of language.
% TODO: How does Go want to archive this?
Go abstracts the creation of threads by using lightweight \emph{goroutines}, which can be created by using the \lstinline{go} keyword followed by a named or anonymous function.
The Go runtime maps these goroutines to normal Kernel threads during execution.
Go encourages programmers to use concurrency by \emph{message passing}, which is assumed to be safer and more convenient to use than \emph{shared memory}.
However, it is also possible to make use of \emph{shared memory}, which gives the developer a lot of freedom to orchestrate the parallel execution of threads.
A study conducted by Tu, Liu, Song and Zhan analyzed popular Go projects and their concurrency bugs.
They came to the conclusion that \emph{message passing} produced even more errors than \emph{shared memory}.
Projects written in Go also tend to have more concurrency than projects written in traditional programming languages such as C.~\cite{tu2019go}
Given all those conclusions it is even more important for Go developers to catch concurrency bugs effectively.
Together with the increasing popularity of the language, the demand for production ready tools to catch concurrency bugs is very high, which is why this paper evaluates existing techniques in the context of the Go programming language.

\subsection{Table of Contents}

\Cref{sct:taxonomy} gives a brief introduction on the different types of concurrency bugs and their main causes: Deadlocks, Data races and Atomicity and Order violations.
\Cref{sct:dynamic} covers some techniques to dynamically detect concurrency bugs, reliably reproduce them by utilizing record and replay with the tool \emph{rr} and detect data races dynamically with the tool \emph{ThreadSanitizer}.
\Cref{sct:testing} shows different methods of concurrency-aware testing to detect concurrency bugs automatically by manipulating the thread scheduler and delta debugging.
And \Cref{sct:static} finally covers how to detect concurrency bugs with static code analysis by sequentialization and model-checking.
In the end there is a conclusion with a short comparison of all techniques and a quick outlook on the possible future of multi-threaded debugging.

@@

\subsection{Deadlocks}
\begin{lstlisting}[float=h, language=Go, label=lst:deadlockWG, caption=Deadlock caused by waiting for the \emph{WaitGroup} at a wrong location -- based on \cite{tu2019go}]
func main() {
    var group sync.WaitGroup
    group.Add(len(data))
    for _, d := range data {
        go func(d string) {
            fmt.Printf("Processing: %s\n", d)
            defer group.Done()
        }(d)
        group.Wait() // Causing the bug
    }
}
\end{lstlisting}

The most commonly manifestation of blocking bugs are \emph{deadlocks}, where circular dependencies between resources block the flow of a program.
\Cref{lst:deadlockWG} shows one example of such a deadlock in a Go program.
The problem is a \emph{blocking synchronization} where the \lstinline{group.Wait()} inside the for-loop is causing the block.
This statement has to be moved outside the for loop to resolve the unintentional blocking and fix the concurrency bug.
Although the error seems obvious in this case, those small mistakes can quickly happen and can get unrecognized into the production environment if not tested well enough.

\begin{lstlisting}[float=h, language=Go, label=lst:deadlockCh, caption={Deadlock caused by misuse of an \emph{unbuffered Channel}}]
func main() {
    ch := make(chan int)
    pr := make(chan int)
    ch <- 1 // Program already blocks here
    go func() {
        i := <-ch
        pr <- i
    }()
    <-pr
}
\end{lstlisting}

A second example of a deadlock that might not be obvious is \Cref{lst:deadlockCh} which uses two unbuffered channels to transfer information between threads.
The problem here is that without an active listener on an unbuffered channel, any send action will be blocked.
To fix this, one could replace the unbuffered channel with a buffered one, so that the execution flow of the program can continue without blocking.
This shows how important it is to know the concrete implementation of a concurrency abstraction.
Even though intended to ease the synchronization between threads and make inter-thread communication safer, by not knowing the implementation of an abstraction the developer can unknowingly create hard to find concurrency bugs.

@@

\subsection{Data Races}
\begin{lstlisting}[float=h, language=Go, label=lst:race, caption=Data race -- based on \cite{goRaceDetector}]
func main() {
	c := make(chan bool)
	m := make(map[string]string)
	go func() {
		m["1"] = "a" // First conflicting access.
		c <- true
	}()
	m["2"] = "b" // Second conflicting access.
	<-c
}
\end{lstlisting}

Data races are part of the non-blocking concurrency bugs.
They occur when multiple threads simultaneously try to access a shared variable where at least one access is a write operation.~\cite{serebry2009threadsanitizer}

\Cref{lst:race} shows an example of a data race that is frequently found.~\cite{serebry2009threadsanitizer}
The data race happens when ``two threads access a non-thread-safe complex object [e.g. a map] without synchronization.''~\cite{serebry2009threadsanitizer}
Even though the two threads in this example write to different keys of the map \lstinline{m}, this might cause a corruption of data or even crash the program because the default Go map implementation is not concurrency-aware.
To fix this, the access to the map needs to be synchronized by a lock for example.

% TODO: Is it a data race or an atomicity violation?
A special case of data races are multi-variable data races.
In this case multiple variables that are semantically correlated are accessed by multiple threads and could therefore loose their semantic correctness when not handled correctly.
A common example for this are structs that have a data and a length field.
Whenever the data field gets updated, so does the length field which is semantically correlated with the data.
When multiple threads update these variables they need to synchronize not only on each of the variables separately but on all semantically correlated variables at the same time.~\cite{lu2007muvi}

@@

Atomicity and order violations are concurrency bugs where the interleaving of threads violate the programmer's intention of atomicity and order.
This often happens because:
``Most programmers think sequentially and therefore they make mistakes easily when writing concurrent programs.''~\cite{lu2008mistakes}

\Cref{lst:order} shows a common order violation bug pattern called ``Test-and-Use''.
The programmer's intention is to check if a variable is not \lstinline{nil} and then use this variable.
However, due to the thread that was launched before, it could happen that after the \lstinline{if} check in line 7, the thread of the goroutine gets scheduled and the data variable is set to \lstinline{nil}.
To fix this bug, the check and the usage of the variable need to become an atomic operation to enforce the order of execution.

\begin{lstlisting}[float=h, language=Go, label=lst:atomicity, caption=Load-Store bug pattern -- Atomicity violation]
func main() {
    var group sync.WaitGroup
    group.Add(10)
    sum := 0
    for i = 0; i < 10; i++ {
        go func() {
            defer group.Done()
            sum++ // Not atomic
        }()
    }
    group.Wait()
}
\end{lstlisting}

\Cref{lst:atomicity} shows one example of the infamous ``Load-Store'' bug pattern.
The programmer assumes that ++ is an atomic operation because it is one literal in Go.
However, after compilation this is expanded to 3 instructions: LOAD, INCREMENT and finally STORE.
The thread scheduler could switch the context after any of these instructions what leads to undefined behavior, when multiple threads try to increment the same variable.
To fix this bug pattern, the \lstinline{++} operation also needs to be replaced by an atomic operation that does not allow other threads to access the \lstinline{sum} variable while incrementing.

Lu, Park, Seo and Zhou conducted a study in 2008 where they analyzed the characteristics of real-world concurrency bugs.~\cite{lu2008mistakes}
One key finding was that:
``Most of the examined non-deadlock concurrency bugs are covered by two simple patterns: atomicity-violation and order-violation''~\cite{lu2008mistakes}

A promising solution to atomicity and order violation bugs is \emph{software transactional memory} (STM) as proposed by Peyton Jones.~\cite{peytonjones2007beautiful}
It is an alternative to traditional lock-based synchronization where atomic regions get declared explicitly.
Languages like Haskell provide STM by their design of language but Go can also utilize this mechanism by using external libraries.

@@

>>>>

@@

\lstdefinestyle{mystyle}{
    backgroundcolor=\color{backcolour},
    commentstyle=\color{codegreen},
    keywordstyle=\color{magenta},
    numberstyle=\tiny\color{codegray},
    stringstyle=\color{codepurple},
    basicstyle=\ttfamily\footnotesize,
    breakatwhitespace=false,
    breaklines=true,
    captionpos=b,
    keepspaces=true,
    numbers=left,
    numbersep=5pt,
    showspaces=false,
    showstringspaces=false,
    showtabs=false,
    tabsize=2,
    frame=single,
    xleftmargin=2em,
    framexleftmargin=1.57em,
    framexrightmargin=-0.5em
}

\lstset{style=mystyle}

@@

% ------------------------------------ %
% ------------- ABSTRACT ------------- %
% ------------------------------------ %
\begin{abstract}
Concurrency bugs are often hard to find and therefore hard to fix.
In this paper we evaluate different techniques and tools that are currently available to effectively find, fix and avoid concurrency bugs.
The requirements for these tools are that they need to be easy to use, they should bring only a minimal computational and storage overhead and they should have a high coverage with minimal false-positive reports.
In short: These tools should be production-ready and applicable to real-world software projects.
The techniques evaluated are:
\begin{itemize}
    \item Dynamic code analysis with record and replay and dynamic data race detection.
    \item Concurrency-aware testing with a combined approach of evaluating thread schedules with delta-debugging to automatically pinpoint concurrency bugs.
    \item Static code analysis with the methods of sequentialization and model-checking.
    \item Language and software design to avoid concurrency bugs
\end{itemize}
\end{abstract}

@@

% Motivation
Today's computer architectures rely more and more on parallel computing.
Single core processors were widely replaced by multi-core architectures optimized to run processes and threads concurrently.
To utilize the full performance of the given hardware, programmers need to write their software with a high degree of parallelism.
In contrast to sequential programs which only have one deterministic path through the program code, concurrent programs have multiple paths of execution that are executed independently from each other.
The order in which these threads are executed is determined by the operating system's scheduler.
The resulting order of thread executions is often referred to as \emph{thread interleavings}.
Because a developer, like the program itself, cannot predict what resources also need processing time on a machine, the concrete schedule has to be handled as non-predictable and therefore as non-deterministic.
This introduces the potential for bugs that would not occur in sequential programs due to non-determinism in the order of execution.
Such bugs can be very challenging to debug because they are not easily reproducible~\cite{tu2019go}.
Due to this complexity, it is no surprise that bugs that are related to concurrency and parallel program execution occur quite often.
But non-deterministic behavior also means that a bug that is only triggered when a certain pattern of thread interleavings appear, cannot be reliably reproduced without knowing the schedule it appeared in.

@@

\begin{quote}
``But Heisenbugs may elude a bugcatcher for years of execution. Indeed, the bugcatcher may perturb the situation just enough to make the Heisenbug disappear.''\cite{gray1986computers}
\end{quote}

Heisenbugs are bugs that only occur with very specific thread interleavings that might not occur when a bugcatcher evaluates the program.
The complexity and uncertainty of concurrency bugs is also the reason for many studies that examine how these bugs occur and explore different ways on how to detect, fix and avoid them~\cite{tu2019go}.
Thus, numerous tools and techniques have been developed that propose different solutions to solve the problem of concurrency bugs and accelerate the debugging process.

@@

\subsection{The Go Programming Language}
The increasing demand for parallel programs has also lead to the development of new programming languages which specialize on this task.
The goal of these languages is to ease the usage of parallel programming patterns so developers can use it more often without writing extensively more code.
Additionally they try to avoid the creation of concurrency bugs by their design of language which will be evaluated in the Outlook section.
One example for such a language is Go which is a statically typed programming language developed by Google, which compiles down to single native binaries.
Go recently gained a lot of popularity for the purpose of writing highly concurrent programs.
Popular open-source projects like \emph{Kubernetes}\footnote{\url{https://kubernetes.io/}}, \emph{Docker}\footnote{\url{https://www.docker.com/}} or \emph{Prometheus}\footnote{\url{https://prometheus.io/}} are all written in Go.
In the official documentation of Go states: ``[... Go's] concurrency mechanisms make it easy to write programs that get the most out of multicore and networked machines [...]''~\cite{goDocs}
Go abstracts the creation of threads by using lightweight \emph{goroutines}, which can be created by using the \lstinline{go} keyword followed by a named or anonymous function.
The Go runtime maps these goroutines to normal Kernel threads during execution.
Go encourages programmers to use concurrency by \emph{message passing}, which is assumed to be safer and more convenient to use than \emph{shared memory}.
For this purpose Go introduces a concept called \emph{channels} where typed messages can be passed through like a pipe.
Beside the message passing however, it is also possible to make use of \emph{shared memory}, which gives the developer a lot of freedom to orchestrate the parallel execution of threads.
Even though Go is designed for programming concurrent programs, this does not mean that Go programs are free of concurrency bugs as we will see in \Cref{sct:outlook}.
Projects written in Go also tend to have more concurrency than projects written in traditional programming languages such as C~\cite{tu2019go}.
So it is even more important for Go developers to catch concurrency bugs effectively and as early as possible.
Together with the increasing popularity of the language, the demand for production ready tools to catch concurrency bugs is very high, which is why this paper evaluates existing techniques in the context of the Go programming language.

\subsection{Structure}

The further structure of this paper continues as follows:
\Cref{sct:taxonomy} gives a brief introduction on the different types of concurrency bugs and their main causes.
\Cref{sct:dynamic} covers some techniques to dynamically detect concurrency bugs, reliably reproduce them.
\Cref{sct:testing} shows different methods of concurrency-aware testing to detect concurrency bugs automatically in a testing environment.
\Cref{sct:static} covers how to detect concurrency bugs with static code analysis.
And finally, \Cref{sct:outlook} will provide a quick outlook on how the examined tools can be used together as well as other promising technologies that might help avoiding concurrency bugs in the future.

@@

\subsection{Deadlocks}

The most common manifestation of blocking bugs are \emph{deadlocks}, where circular dependencies between resources block the flow of a program.

\begin{lstlisting}[float=h, language=Go, label=lst:deadlockCh, caption={Deadlock caused by misuse of an \emph{unbuffered Channel}}]
func main() {
    data := make(chan int)
    done := make(chan bool)
    ch <- 1 // Program already blocks here
    go func() {
        process(<-data)
        done <- true
    }()
    <-done
}
\end{lstlisting}

An example of a deadlock in a Go program that might not be obvious is \Cref{lst:deadlockCh} which uses unbuffered channels to transfer information between threads.
Go channels can either be buffered or unbuffered.
With unbuffered channels the sender thread has to wait until a second thread is ready to receive the message.
With the buffered channel the message gets buffered and the sender thread can continue it's execution immediately as long as the buffer is not full.
In the first two lines of the main method the two unbuffered channels \lstinline{data} and \lstinline{done} are created to send data to the thread and to receive information when the execution of the thread is done.
Using the arrow notation data can be either send into a channel (line 4) or data can be received from a channel (line 6)
The bug in this case is that without an active listener on an unbuffered channel, any send action will be blocked.
To fix this, one could replace the unbuffered channel with a buffered one, so that the execution flow of the program can continue without blocking.
This shows how important it is to know the concrete implementation of a concurrency abstraction.
Even though intended to ease the synchronization between threads and make inter-thread communication safer, by not knowing the implementation of an abstraction the developer can unknowingly create hard to find concurrency bugs.

@@

\subsection{Data Races}

Data races are part of the non-blocking concurrency bugs.
They occur when multiple threads simultaneously try to access a shared variable where at least one access is a write operation~\cite{serebry2009threadsanitizer}.

\begin{lstlisting}[float=h, language=Go, label=lst:race, caption=Data race by concurrently accessing a map]
func main() {
	data := make(map[string]string)
	done := make(chan bool)
	go func() {
		data["1"] = "a" // First conflicting access.
		done <- true
	}()
	data["2"] = "b" // Second conflicting access.
	<-done
}
\end{lstlisting}

\Cref{lst:race} shows an example of a data race that is frequently found.~\cite{serebry2009threadsanitizer}
The data race happens when two threads access a non-thread-safe complex object [e.g. a map] without synchronization~\cite{serebry2009threadsanitizer}.
Even though the two threads in this example write to different keys of the map \lstinline{data}, this might cause a corruption of data or even crash the program because the default Go map implementation is not concurrency-aware.
To fix this, the access to the map needs to be synchronized, e.g. by a lock.

% TODO: Is it a data race or an atomicity violation?
A special case of data races are multi-variable data races.
In this case multiple variables that are semantically correlated are accessed by multiple threads and could therefore loose their semantic correctness when not handled correctly.
A common example for this are structs that have a data and a length field.
Whenever the data field gets updated, so does the length field which is semantically correlated with the data.
When multiple threads update these variables they need to synchronize not only on each of the variables separately but on all semantically correlated variables at the same time~\cite{lu2007muvi}.

@@

Atomicity and order violations are concurrency bugs where the interleaving of threads violates the programmer's intention of atomicity and order.
% TODO: cite test and use
\Cref{lst:order} shows a common order violation bug pattern called ``Test-and-Use''.
The programmer's intention is to check if a variable is not \lstinline{nil} and then use this variable.
However, due to the thread that was launched before, it could happen that after the \lstinline{if} check in line 7, the thread of the goroutine gets scheduled and the data variable is set to \lstinline{nil}.
To fix this bug, the check and the usage of the variable need to become an atomic operation to enforce the order of execution.

\begin{lstlisting}[float=h, language=Go, label=lst:atomicity, caption=Load-Store bug pattern -- Atomicity violation]
func main() {
    var group sync.WaitGroup
    group.Add(10)
    sum := 0
    for i = 0; i < 10; i++ {
        go func() {
            sum++ // Not atomic
            group.Done()
        }()
    }
    group.Wait()
}
\end{lstlisting}

% TODO: cite load-store
\Cref{lst:atomicity} shows one example of the infamous ``Load-Store'' bug pattern.
The programmer might have assumed that \lstinline{++} is an atomic operation because it is one literal in Go.
However, after compilation, this is expanded to 3 instructions: LOAD, INCREMENT and finally STORE.
The thread scheduler could switch the context after any of these instructions.
This leads to undefined behavior, when multiple threads try to increment the same variable.
To fix this bug pattern, the \lstinline{++} operation also needs to be replaced by an atomic operation that does not allow other threads to access the \lstinline{sum} variable while incrementing.

@@

% ------------------------------------ %
% -------------- OUTLOOK ------------- %
% ------------------------------------ %
\section{Outlook}
\label{sct:outlook}
A study conducted by Tu, Liu, Song and Zhan analyzed popular Go projects and their concurrency bugs.
They came to the conclusion that \emph{message passing} produced even more errors than \emph{shared memory}.

Lu, Park, Seo and Zhou conducted a study in 2008 where they analyzed the characteristics of real-world concurrency bugs~\cite{lu2008mistakes}.
One key finding was that:
``Most of the examined non-deadlock concurrency bugs are covered by two simple patterns: atomicity-violation and order-violation''~\cite{lu2008mistakes}

A promising solution to atomicity and order violation bugs is \emph{software transactional memory} (STM) as proposed by Peyton Jones~\cite{peytonjones2007beautiful}.
It is an alternative to traditional lock-based synchronization where atomic regions get declared explicitly.
Languages like Haskell provide STM by their design of language but Go can also utilize this mechanism by using external libraries.
